---
title: "Comparisons"
---

```{r include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(chorddiag)
library(htmlwidgets)
library(igraph)
library(readr)
library(tidygraph)
library(tidyverse)
library(DT)

```

## Comparing Past vs. Present Conditions

Using LANDFIRE’s BpS products, we explore two different ways to visualize past vs. current vegetation patterns.

-   First, we present **changes in broad ecosystem types** using an interactive comparison diagram. To present these broad ecosystem trends more effectively, we combined classifications of Existing Vegetation Types (EVT) into broader categories. 

-   Second, we compare **amounts of succession classes** (past and present) for the most prevalent ecosystems.



*Note: number presented when hovering equals acres.*

**PAST** &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp; &emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp; &emsp;&emsp; &emsp; **CURRENT**

```{r chord, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# read in data
chord_df<- read_csv("data/bps2evt_chord2.csv")
#view(histFireGVchord)

#convert to matrix
matrix_df <-as.matrix(as_adjacency_matrix(as_tbl_graph(chord_df),attr = "ACRES"))

#clean up matrix (could be cleaner!)
matrix_df = subset(matrix_df, select = -c(1:9))

matrix_df <- matrix_df[-c(10:15),]

#make a custom color pallet #eb4034 (redish) #b0af9e(grey)

# ORIGINAL
groupColors <-c( "#f5e942", # cur ag
                 "#1d4220", # cur conifer
                 "#397d3f", # cur hdw-con
                 "#b0af9e", # developed
                 "#eb4034", # exotics
                 "#fc9d03", # grassland
                 "#56bf5f", # hardwood
                 "#7db7c7",
                 "#6e4f1e", #  shrubland
                 "#1d4220", # conifer 
                 "#fc9d03", # grassland
                 "#56bf5f", # hardwood
                 "#397d3f", # hardwood-conifer 
                 "#7db7c7", # riparian 
                 "#6e4f1e" # shrubland

                 

)






#make chord diagram
chord<-chorddiag(data = matrix_df,
                 type = "bipartite",
                 groupColors = groupColors,
                 groupnamePadding = 10,
                 groupPadding = 3,
                 groupnameFontsize = 11 ,
                 showTicks = FALSE,
                 margin=120,
                 tooltipGroupConnector = "    &#x25B6;    ",
                 chordedgeColor = "#363533"
)
chord 

#save then print to have white background
htmlwidgets::saveWidget(chord,
                        "chord.html",
                        background = "white",
                        selfcontained = TRUE
                        )
```



<iframe src="chord.html" height="720" width="720" style="border: 1px solid #464646;" allowfullscreen="" allow="autoplay" data-external=".5"></iframe>

<br>

## Succession classes for most dominant Biophysical Settings





```{r scls chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=9}

scls_descriptions <- read_csv("data/scls_descriptions.csv") |>
  unite(model_label, c("StratumID", "ClassLabelID"), remove = FALSE)

BPS_SCLS2 <- read.csv("data/bpsScls2.csv") 

bps_scls_3 <- BPS_SCLS2 %>%
  group_by(Var1) %>%
  mutate(total.count = sum(Freq)) %>%
  ungroup() %>%
  dplyr::filter(dense_rank(desc(total.count)) < 8) %>%
  dplyr::select(c("bps_name", "ref_label",  "currentPercent", "ref_percent", "model_label")) %>%
  pivot_longer(
    cols = c(`ref_percent`, `currentPercent`), 
    names_to = "refCur", 
    values_to = "Percent"
    ) |>
  left_join(scls_descriptions)




# order classes
bps_scls_3$ref_label <- factor(bps_scls_3$ref_label, levels= c(
  "Developed",
  "Agriculture",
  "UE",
  "UN",
  "E",
  "D",
  "C",
  "B",
  "A"))




sclasplot <-
  ggplot(bps_scls_3, aes(fill=factor(refCur), y=Percent, x=ref_label)) + 
  geom_col(width = 0.8, position = position_dodge()) +
  coord_flip() +
  facet_grid(. ~BpS) +
  scale_x_discrete(limits = (levels(bps_scls_3$ref_label))) +
  labs(
    title = "Succession Classes past and present",
    subtitle = "Top BpSs selected for illustration. Not all succession classes present in all BpSs",
    caption = "\nData from landfire.gov.",
    x = "",
    y = "Percent")+
  theme_minimal(base_size = 12)+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") +
  scale_fill_manual(values = c("#3d4740", "#32a852" ), # present (grey), historical (green)
                    name = " ", 
                    labels = c("Present",
                               "Past")) +
  facet_wrap(~bps_name, nrow(3),labeller = labeller(bps_name = label_wrap_gen())) +
    theme(panel.spacing = unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1))

sclasplot

```

## Succession Class Descriptions

Below is a table of the Succession Class descriptions for the Biophysical Settings charted above.  For complete Biophysical Settings descriptions see links on Past page. 


```{r}
#| label: sclass descriptions
#| echo: false
#| message: false
#| warning: false


for_scl_table <-  bps_scls_3 |>
  filter(across(everything(), ~ !is.na(.x))) |>
  filter(str_detect(refCur, "ref")) |>
  select(c(bps_name,
           ref_label,
           StateClassID,
           Description
           )) |>
  distinct(Description, .keep_all = TRUE) |>
  rename(c('Biophysical Setting' = bps_name,
           'Class' = ref_label,
           'Class Label' = StateClassID,
           ))


# Render the table with DT
datatable(for_scl_table, options = list(
  columnDefs = list(list(
    targets = 3, # Index of the Description column (0-based)
    render = JS("function(data, type, row, meta) {
      return type === 'display' && data.length > 50 ? 
        '<span title=\"' + data + '\">' + data.substr(0, 50) + '...</span>' : data;
    }")
  )),
  #scrollX = TRUE,
  autoWidth = TRUE
))





```






